Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-14 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
section 3.2, 3.3.2

A partial answer to these questions is provided in \cite{kadkhodaie2024generalization} within the context of diffusion-based models, highlighting the impact of dataset size on the transition from \textit{memorization} to \textit{generation} regimes. Following their methodology, we compare two generative models of the same architecture trained on non-overlapping subsets of the same dataset. For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights.

All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
%

@@

@@

The distributions of morphological coefficients derived from the three models show good agreement with those computed from the original images. However, we do not aim for high-quality images or state-of-the-art results, as seen in \citep{ravanbakhsh2016, Fussell2019, Lanusse2021, smith2021, HACKSTEIN2023100685}, which achieve better image quality. Instead, our focus is on a different question: with low-resolution and relatively simple images, do models of the same type trained on moderately sized datasets learn the same underlying probability distribution? Furthermore, if differences exist, can we identify measures to assess the reliability of the results?
%
\subsubsection{\texttt{U-Net} Denoising Performances}
\label{sec-UNet-Perf}
%
Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.

@@

This disparity in denoising performance between the two networks has significant implications for the generative process, which begins with a purely noisy image and progressively refines it into a noiseless output.
%
\subsubsection{Two Diffusion-Based Models Test}
\label{sec-two_models-Diffusion}
%
Figure~\ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric \citep{books/aw/TanSK2005}, considering two distinct scenarios. The first scenario (top plots) involves $1,000$ samples generated by two diffusion-based models ("Model A" and "Model B") trained on independent datasets of the same size $N$, with generation \textit{seeded by the same white noise images}. The second scenario (bottom plots) examines a single model, comparing each of its $1,000$ generated samples to the closest training image in terms of cosine similarity.

>>>>

A partial answer to these questions is provided in \cite{kadkhodaie2024generalization} within the context of diffusion-based models, highlighting the impact of dataset size on the transition from \textit{memorization} to \textit{generation} regimes. Following their methodology, we compare two generative models of the same architecture trained on non-overlapping subsets of the same dataset. For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights. 
{\color{red}We focus on investigating how one can perform tests, particularly with the two-model scenario, to ensure that the model behaves as intended. 
}

%14-mars-25 phrase deplacee
%All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
%

@@

{\color{red}The author have perform some tests with other settings of the above architectures and even if not exhaustive, the results presented in the following section are generic and would serve as guide line if one would like to use either another settings of the models I have used, either to test a completely different model architecture.

All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
}

@@

The distributions of morphological coefficients derived from the three models show good agreement with those computed from the original images. However, we do not aim for high-quality images or state-of-the-art results, as seen in \citep{ravanbakhsh2016, Fussell2019, Lanusse2021, smith2021, HACKSTEIN2023100685}, which achieve better image quality. Instead, our focus is on a different question: with low-resolution and relatively simple images, do models of the same type trained on moderately sized datasets learn the same underlying probability distribution? Furthermore, if differences exist, can we identify measures to assess the reliability of the results?
%
\subsubsection{\texttt{U-Net} Denoising Performances}
\label{sec-UNet-Perf}
%
%Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.
{\color{red} Before assessing the generative performance of the diffusion models in the next section, we first present the denoising performance of two \texttt{U-Net} networks with the same architecture, which are subsequently used to estimate the score as discussed in Section~\ref{sec-diff} (Equation~\ref{eq-backward-diffusion}).
}

@@

{\color{red}This disparity in denoising performance between the two networks may have significant implications for the generative process, as the score computation could be problematic, starting with a purely noisy image and progressively refining it into a noiseless output.}
%
\subsubsection{Two Diffusion-Based Models Test}
\label{sec-two_models-Diffusion}
%
Figure~\ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric \citep{books/aw/TanSK2005}, considering two distinct scenarios. The first scenario (top plots) involves $1,000$ samples generated by two diffusion-based models ("Model A" and "Model B") trained on independent datasets of the same size $N$, with generation \textit{seeded by the same white noise images}. The second scenario (bottom plots) examines a single model, comparing each of its $1,000$ generated samples to the closest training image in terms of cosine similarity.
